---
title: "Semester_Project"
author: "Rob McNeil"
date: "2024-11-26"
output: html_document
---

#assess type_ and facility_type for missingness
#aggregate: create subgroups so you have 2 or three types
#how do I handle types that use both bup and nax? For geospatial, you can decide what you want to investiate
#can make 4 groups, no, 1, other, both, can compare in terms of how they are spatially distributed
#go to the hackathon meetings on monday
#base R subsetting 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r, message=FALSE, warning=FALSE}

library(spatstat)
library(ggmap)
library(ggplot2)
library(knitr)
SAMSHA <-read.csv("C:/Users/rrm10/Downloads/SAMHSA.csv")



```

```{r,warning=FALSE, echo=FALSE}

#str(SAMSHA)


#explore missingness between Type_ and Facility_Type 

unique(SAMSHA$Type_)
unique(SAMSHA$Facility_Type)

length(SAMSHA$Type_)
length(SAMSHA$Facility_Type)

sum(is.na(SAMSHA$Type_))
sum(is.na(SAMSHA$Facility_Type))

#neither column has missingness, I think I'm going to use Facility_Type because it will be easier for me to stratify

```




```{r, warning=FALSE,message=FALSE}

#provide range in how we want to consider obervations
bbx<- owin(xrange = range(SAMSHA$LONGITUDE), yrange = range(SAMSHA$LATITUDE))
#planar point pattern: specify 2 coloumns in dataset that represent geo-spatial dimensions, store it in object X
X <- ppp(SAMSHA$LONGITUDE, SAMSHA$LATITUDE, window = bbx)
#warning (data contains duplicated points)
X
plot(X)
# create basic visualization, can see density follows front range

# Get the number of rows in SAMSHA dataset
n <- dim(SAMSHA)[1]  # Get number of rows (observations)

# Initialize a vector to store facility type marks, initially set to 0
fmv <- rep(0, n)

#use regular expression to stratify mental health facility 
tmp <-grep("psychiatric|mental",SAMSHA$Facility_Type, ignore.case=TRUE)


fmv[tmp]<-1

fmv1<- factor(fmv,levels = c(0,1), labels = c("General Facility", "Psychiatric Facility"))

X.m <- ppp(SAMSHA$LONGITUDE,SAMSHA$LATITUDE,window=bbx,marks = fmv1)
plot(X.m)

tmp<-grep("residential", SAMSHA$Facility_Type, ignore.case=TRUE)
fmv[tmp]<-2
fmv2 <-factor(fmv)
X.m2 <- ppp(SAMSHA$LONGITUDE,SAMSHA$LATITUDE,window=bbx,marks=fmv2)
plot(X.m2)




```

```{r}

# Perform quadrat counting with 8x8 bins
Q8 <- quadratcount(X.m2, nx=8, ny=8)

plot(Q8,main= " Table 1: Quadrat Count Plot")

# Perform the quadrat test for the observed point pattern X.m (you may also want to test X.m2)
quadrat_test_result <- quadrat.test(X.m2, nx=8, ny=8)

# Print the test result
print(quadrat_test_result)



```

Statistical Inference using K function 
```{r, warning=FALSE}
estK <-Kest(X)
plot(estK)
#question at 20 minutes should I be making seperate vectors for each>
#this method doesnt consider marks, only considers distrobution of datapints (27:00)
#null hypothesis assumes a complete spatial randomness
#why do I get 4 lines? 

swb<-rescale(X)
kvb<-varblock(swb,Kest,nx=8,ny=8) #bootstrap function utilizing binning 
plot(kvb)

kvl <-lohboot(swb,Kest)    #confidence intervals, runs many simulations. more sophisticated
plot(kvl)


E <- envelope(swb,Kest)  #summary of simulation
E




```
other method based on stratification of different marks at 44:00

```{r, warning=FALSE}

probH <-relrisk(X.m2,diggle=T) #non-parametric test estimating relative risk
plot(probH, main="Density Plot (by Facility Type)")
#consider increasing to 100 or 200 for semester project
seg_test_fac <-segregation.test(X.m2,nsim=200,hmin=0.05,hmax=5)



```

```{r}
#make table for writeup output 

# Extract values from the segregation test result
test_statistic <- seg_test_fac$statistic
p_value <- seg_test_fac$p.value
method <- seg_test_fac$method
data_name <- seg_test_fac$data.name

# Create a data frame to structure the output
segregation_result <- data.frame(
  Metric = c("Test Statistic", "p-value", "Method", "Data Name"),
  Value = c(test_statistic, p_value, method, data_name)
)

# Use kable to create a nicely formatted table

kable(segregation_result, caption = "Table 1: Segregation Test Results (By Facility)", format = "pipe")




```




Data Visualization 


```{r}



register_stadiamaps(key = "93ad8b2e-920b-4037-8f94-97ec41ee397d")

# Step 1: Extract coordinates and facility types from X.m2
coords_df <- data.frame(
  Longitude = X.m2$x,  # Longitude values from X.m2
  Latitude = X.m2$y,   # Latitude values from X.m2
  Facility_Type = factor(X.m2$marks, levels = c("0", "1", "2"), 
                         labels = c("General Facility", "Psychiatric Facility", "Residential Facility"))
)

# Step 2: Get the basemap (adjust zoom if needed)
colo <- c(left = -109, bottom = 37, right = -102, top = 41)
colo_map <- get_stadiamap(colo, zoom = 7, maptype = "stamen_terrain") |> ggmap()

# Step 3: Create the map with ggplot2, overlaying the points
colo_map + 
  geom_point(aes(x = Longitude, y = Latitude, color = Facility_Type), data = coords_df, size = 3) +
  scale_color_manual(values = c("General Facility" = "blue", "Psychiatric Facility" = "red", "Residential Facility" = "green")) +
  theme_minimal() +
  labs(title = "Mental Health Facility Distribution",
       subtitle = "Facilities marked by type (General, Psychiatric, Residential)",
       x = "Longitude", y = "Latitude")






```




Question 2: Are spatial distribuions of behavioral health treatment service providers throughout Colorado different based on whether Naltrexone or buprenorphine is used in treatment> 

```{r,echo=FALSE,results='hide'}

#exploring Opiod_Treatment_Settings 

unique(SAMSHA$Opiod_Treatment_Settings)
#Can see Naltrexone used in treatment, does not treat opiod additction 

table(SAMSHA$Opiod_Treatment_Settings, SAMSHA$Naltrexone_used_in_Treatment, SAMSHA$Buprenorphine_used_in_Treatment)

```





```{r, warning=FALSE}

# Initialize a vector to store treatment type marks, initially set to 0
amb <- rep(0, n)

#use regular expression to stratify mental health facility 
bmp <-grep("Naltrexone used in Treatment",SAMSHA$Opiod_Treatment_Settings, ignore.case=TRUE)
cmp <-grep("Buprenorphine used in Treatment", SAMSHA$Opiod_Treatment_Settings,ignore.case=TRUE)
#gmp <-grep("Buprenorphine&Naltrexone", SAMSHA$Opiod_Treatment_Settings,ignore.case=TRUE)

#regex for "false positives" 

#dmp<-grep("Naltrexone used in Treatment.*(Does not treat opioid addiction|Do not use medication for opioid addiction)", SAMSHA$Opiod_Treatment_Settings,ignore.case=TRUE)

#emp <-grep("Buprenorphine used in Treatment.*(Does not treat opioid addiction|Do not use medication for opioid addiction)",
           #SAMSHA$Opiod_Treatment_Settings, ignore.case = TRUE)

amb[bmp]<-1
amb[cmp]<-2
#amb[gmp]<-3
#amb[dmp]<-0
#amb[emp]<-0

ymv <-factor(amb, levels = c(0, 1, 2), labels = c("No Medication", "Naltrexone", "Buprenorphine"))


Y.m <- ppp(SAMSHA$LONGITUDE, SAMSHA$LATITUDE, window = bbx, marks = ymv)



```



```{r, warning =FALSE}

#statistical inference
probI <- relrisk(Y.m, diggle = TRUE)  # Non-parametric test estimating relative risk
plot(probI, main="Density Plot (by treatment")

seg_treatment <-segregation.test(Y.m, nsim = 200, hmin = 0.05, hmax = 5)



```

```{r}
# Vary the bandwidth to see its impact on the relative risk estimate
bandwidth_values <- c(0.5, 1, 2, 5)  # bandwidth increased to avoid underflow 

# Loop over different bandwidth values and calculate relative risk for each
for (bw in bandwidth_values) {
  probI <- relrisk(Y.m, diggle = TRUE, sigma = bw)  # Adjust bandwidth
  plot(probI, main = paste("Density Plot (Bandwidth =", bw, ")"))
}


```
Making a table from segregation test
```{r}
# Extract values from the segregation test result
test_statistic <- seg_treatment$statistic
p_value <- seg_treatment$p.value
method <- seg_treatment$method
data_name <- seg_treatment$data.name

# Create a data frame to structure the output
segregation_results <- data.frame(
  Metric = c("Test Statistic", "p-value", "Method", "Data Name"),
  Value = c(test_statistic, p_value, method, data_name)
)

# Use kable to create a nicely formatted table

kable(segregation_results, caption = "Table 2: Segregation Test Results (By Treatment)", format = "pipe")



```


Data Visualization Using Smoothing clusters 
To understand the concentratio of facilities, density plot will highlight regions with higher densicy 


Statistical inference
Stratified Point Pattern: Since Y.m has categorical marks (i.e., treatment types), you can perform the relative risk and segregation test in a stratified manner, as you've done with X.m2. In Y.m, the marks (Naltrexone, Buprenorphine, No Medication) represent different treatments, so you are comparing the spatial patterns of these three treatment types.





```{r}

#  Extract coordinates and treatment from ppp Y.m
coords_df_y <- data.frame(
  Longitude = Y.m$x,  
  Latitude = Y.m$y,   
   treatment_type = factor(Y.m$marks, levels = c("No Medication", "Naltrexone", "Buprenorphine"))
)


# Background map with density overlay
colo_map_background <- get_stadiamap(colo, zoom = 7, maptype = "stamen_toner_background") |> ggmap(darken = 0.7)


# Overlay density plots for each treatment type with treatment categories as colors
colo_map_background +
  geom_point(data = coords_df_y, 
             aes(x = Longitude, y = Latitude, color = treatment_type), 
             size = 3, alpha = 0.5) +
  scale_color_manual(values = c("No Medication" = "blue", 
                                "Naltrexone" = "red", 
                                "Buprenorphine" = "green")) +
  theme_minimal() +
  labs(title = "Mental Health Facility Distribution by Treatment Type",
       subtitle = "Colored by Treatment Type (Naltrexone, Buprenorphine, No Medication)",
       x = "Longitude", y = "Latitude")


```
